Concurrent Programming in Harmony

### 2021-07-11 book/paper.tex
Writing the consensus chapter

\begin{quote}
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \textit{count} will always end up being 2?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying machine architecture are atomic (indivisible)---in fact they are.)
\end{itemize}
\end{quote}

@@

Harmony programs are compiled to Harmony \emph{bytecode}
\index{bytecode}%
(a list of machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
To understand the problem of concurrent computing, it
is important to have a basic understanding of machine instructions,
and in our case those of the HVM.

@@

\begin{figure}
\begin{code}
\harmonysource{bosco}
\end{code}
\caption{\harmonylink{code/bosco.hny} A simple fault-tolerant consensus protocol.}
\label{fig:bosco}
\end{figure}

>>>>

\begin{quote}
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \textit{count} will always end up being 2?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying virtual machine architecture are atomic (indivisible)---in fact
they are.)
\end{itemize}
\end{quote}

@@

Harmony programs are compiled to Harmony \emph{bytecode}
\index{bytecode}%
(a list of machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
The Harmony compiler places the bytecode for file $x$.\texttt{htm} in
file $x$.\texttt{hvm}.
The model checker, called \emph{Charm}, places its output in a file
called $x$.\texttt{hco} and a stylized version of the same output
in an HTML file called $x$.\texttt{htm}.

To understand the problem of concurrent computing, it
is important to have a basic understanding of machine instructions,
and in our case those of the HVM.

@@

\begin{figure}
\begin{code}
\harmonysource{bosco}
\end{code}
\caption{\harmonylink{code/bosco.hny} A crash-tolerant consensus protocol.}
\label{fig:bosco}
\end{figure}

Distributed consensus is the problem of having a collection of processors agree
on a single value over a network.  This often has to be done in adverse circumstances,
for example in the face of processor failures.
In Harmony, processors can be modeled by threads and there are a variety of ways in
which one can model a network using shared variables.
In this chapter, we will present a simple consensus algorithm that can tolerate
fewer than $1/3^{rd}$ of processors failing by crashing.

More precisely, we will
define a constant \texttt{F} that contains the maximum number of failures, and
assume there are $\texttt{N} = 3\texttt{F} + 1$ processors.
Each processor \emph{proposes} a value, which we assume here is from the
set \{ 0, 1 \}.
By the definition of consensus, we want the following two propertiew:
\begin{enumerate}
\item Validity: a processor can only decide a value that has been proposed;
\item Agreement: if two processors decide, they decide the same value.
\end{enumerate}
The consensus problem is surprisingly tricky to solve in the face of processor
failures.

We will model the network as a bag (aka multiset) of messages.  The bag itself is
modeled in Harmony as a dictionary that maps unique elements in the bag to their
multiplicity.  So, for example, $\{ .\texttt{a}:1, .\texttt{b}:2 \}$ is the bag that consists of
one $.\texttt{a}$ and two copies of $.\texttt{b}$.
The \texttt{bag} module (\autoref{ap:bag}) contains a variety of handy functions on bags.
In this particular algorithm, all messages are broadcast to all processors, so they do
not require a destination address.

\autoref{fig:bosco} presents our algorithm.  Besides the network variable, we will use a
shared list of proposals and a shared set of decisions.  The invariants we want to maintain
on those variables is that $\textit{decisions} \subseteq \textit{proposals}$
and $|\textit{decisions}| \leq 1$.  Since the initial value of \textit{decisions} is
$\emptyset$, the invariants clearly hold initially.

The $N$ processors after go through a sequence of \emph{rounds} in which they wait for
$\texttt{N} - \texttt{F}$ messages, update their state based on the messages, and
broadcast a new messages containing their new state.
The reason that a processor waits for $\texttt{N} - \texttt{F}$ rather than \texttt{N}
messages is because of failures: up to $\texttt{F}$ processors may never send a message
and so it would be unwise to wait for all $\texttt{N}$.  You might be tempted to use
a timer and time out on waiting for a particular processor.  But how would you initialize
that timer?  While we will assume that the network is reliable, there is no guarantee
on that messages arrive within a particular time.
We call a set of $\texttt{N} - \texttt{F}$ processors a \emph{quorum}.
A quorum must suffice for the algorithm to make progress.

The state of a processor consists
of its current round number and proposal.  Therefore, messages consist of a tuple containing
a round number and a proposal. To start things, each processor first
broadcasts its initial round number (0) and proposal.
The number of rounds that are necessary to achieve consensus is not bounded.  But
Harmony can only check finite models, so there is a constant \texttt{NROUNDS} that
limits the number of rounds.  It may be that no decisions are made, but that does not
violate either Validity or Agreement.  We only check that \emph{if} decisions are
made, they satisfy Validity and Agreement.

\index{select statement}%
In Line~21, a processor waits for $\texttt{N} - \texttt{F}$ messages using
the Harmony \textbf{select} statement.
This statement takes the form \textbf{select} $v$ \textbf{in} $s$: \emph{statement block}
where $s$ is a set and $v$ is variable that is bound to an element of $s$.
The important properties of a \textbf{select} statement are as follows:
\begin{itemize}
\item it waits until $s$ is non-empty;
\item it is executed atomically;
\item $v$ is selected non-deterministically, like in the \texttt{choose} operator.
\end{itemize}

Since Harmony has to check all possible executions of the protocol, the
\texttt{receive}(\textit{round}, $k$) method returns all \emph{subbags} of messages
for the given round that have size $k = \texttt{N} - \texttt{F}$.  The method uses a
dictionary comprehension to filter out all messages for the given \textit{round}
and then uses the \texttt{bag.combinations} method to find all combinations of size $k$.
The \textbf{select} statement waits until there is at least one such combination and
then chooses an element that is bound to the \textit{quorum} variable.
The body of the \textbf{select} statement is then executed atomically.
This is usually how distributed algorithms are modeled, because they can only interact
through the network.  There is no need to interleave the different processes other
than when messages are delivered.  By executing the body atomically, a lot of
unnecessary interleavings are avoided and this reduces the state space that must
be explored by the model checker significantly.

The body of the \textbf{select} statement contains the core of the algorithm.
Note that $\texttt{N} - \texttt{F} = 2\texttt{F} + 1$, so that the number of
messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
must exist a majority of zeroes or ones.  Variable \textit{count}[0] stores the number
of zeroes and \texttt{count}[1] stores the number of ones received in the round.
The rules of the algorithm are simple:
\begin{itemize}
\item update \textit{proposal} to be the majority value;
\item if the quorum was unanimous, decide the value.
\end{itemize}
After that, proceed with the next round.
